% ============================================================================
% BYLAWS DOCUMENT - Main editable file
% ============================================================================
% To edit: Update \Article and \Sec sections with your content
% To rebrand: Modify the "Branding / quick tweaks" section (lines 16-23)
% To change colors: Edit the \definecolor lines (hex color codes)
% ============================================================================

\documentclass[11pt]{article}

% Layout & typography
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{lmodern}

% Colors, links, headings, header/footer
\usepackage[dvipsnames]{xcolor}
\usepackage[hidelinks]{hyperref}
\usepackage{titlesec}
\usepackage{fancyhdr}

% ---------- Branding / quick tweaks ----------
% EDIT: Update these values for your organization
\newcommand{\depname}{ECSE}
\newcommand{\orgname}{ECSE Graduate Council}
\newcommand{\orgshort}{ECSE GC}
\newcommand{\doctitle}{\orgname\ Bylaws}
\newcommand{\docversion}{Version 1.0}
\definecolor{Heading}{HTML}{D6001C}  % RPI red
\definecolor{Rule}{HTML}{B22222}

% ---------- Fancy header / footer ----------
\pagestyle{fancy}
\fancyhf{}
\lhead{\textsc{\orgname}}
\rhead{\textsc{Bylaws}}
\cfoot{\thepage}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\headrule}{\hbox to\headwidth{%
  \color{Rule}\leaders\hrule height \headrulewidth\hfill}}
\renewcommand{\footrulewidth}{0pt}

% ---------- Elegant section headings ----------
% Section (Articles)
\titleformat{\section}
  {\Large\bfseries\color{Heading}}
  {}{0pt}{\MakeUppercase}

% Subsection (Sections within Articles)
\titleformat{\subsection}
  {\normalsize\bfseries\color{Heading}}
  {}{0pt}{\scshape}


% Reduce spacing before/after headings a touch
\titlespacing*\section{0pt}{1.0em}{0.75em}
\titlespacing*\subsection{0pt}{0.8em}{0.6em}

% ---------- Utility macros ----------
\newcommand{\Article}[2]{\section*{Article #1. #2}}
\newcommand{\Sec}[1]{\subsection*{Section {#1}}}
\newcommand{\titlerulecustom}{%
  {\color{Rule}\rule{\textwidth}{1.2pt}}%
}

\begin{document}

% ---------- Title block ----------
\begin{center}
  {\huge\bfseries\color{Heading}\doctitle\par}
  \vspace{0.5em}
  {\normalsize Effective: November 19, 2025 \par}
  \vspace{-0.2em}
  \titlerulecustom
\end{center}

% ---------- Body ----------
% Structure: \Article{ROMAN}{Title} for sections, \Sec{number. Title} for subsections
% List items use enumerate environment with [label=\alph*)] for a), b), c)
\Article{I}{Name}
The name of this organization shall be the \textbf{\orgname} (hereafter, the \textbf{\orgshort}).

\Article{II}{Purpose}
The \orgshort{} exists to improve the graduate student experience within the Department of Electrical, Computer, and Systems Engineering (ECSE) by (i) building community among graduate students, (ii) advocating for graduate student needs to departmental and Institute leadership, (iii) organizing social, professional, and wellness programming, and (iv) supporting initiatives that enhance learning, research, and everyday life for \depname{} graduate students.

\Article{III}{Membership}

\Sec{1. Equal Opportunity and Hazing}
\begin{enumerate}[label=\alph*)]
    \item It is the policy of the \orgshort{} to provide, to the maximum extent feasible, equal opportunity to all qualified individuals without regard to race, ethnicity, color, religion, national origin, gender, sexual orientation, or disability for all aspects of the \orgshort{}'s activities.
    \item The \orgshort{} shall make reasonable efforts to ensure Masters', Co-Term and PhD students of the department are adequately represented.
    \item Hazing, reckless or intentional endangerment to health, or forced consumption of alcohol or other drugs for the purpose of initiation into or continuing membership of the \orgshort{} is strictly prohibited.
\end{enumerate}

\Sec{2. Definition of Membership}
\begin{enumerate}[label=\alph*)]
    \item \textbf{General Members} consist of all registered Rensselaer Polytechnic Institute (RPI) students who are currently enrolled in graduate programs offered by the \depname{} Department. General Membership is automatic upon enrollment.
    \item \textbf{Council Members} are General Members who have been elected to hold an office. They serve as the primary voting body for day-to-day \orgshort{} operations subject to a maximum of eight (8).
    \item \textbf{Associate Members} (non-voting) may include but not limited to \depname{} faculty/staff, \depname{} alumni, and non-\depname{} RPI students interested in supporting \orgshort{} activities.
\end{enumerate}

\Sec{3. Obtaining Membership}
\begin{enumerate}[label=\alph*)]
    \item To obtain Council Membership, an General Member must be nominated by any active graduate student in the \depname{} department and must be voted in during regular annual elections as outlined in Section 5.
    \item Associate Membership may be granted through approval by the \orgshort{} upon receipt of written request indicating such intent.
\end{enumerate}

\Sec{4. Continued Membership}
Every regular member of \orgshort{} voted in during regular annual elections shall continue to hold office for 1 year or until the next elections, as the case may be.

\Sec{5. Voting Rights}
\begin{enumerate}[label=\alph*)]
    \item \textbf{General Members:} Have the right to cast exactly one (1) vote in the annual election of Council Members.
    \item \textbf{Council Members:} Have the right to cast exactly one (1) vote on internal Council business, officer appointments, and amendments to these bylaws.
    \item A member loses voting rights upon loss of Regular Member status or for cause as determined through the disciplinary process in Section~7.
\end{enumerate}

\Sec{6. Membership Information}
\begin{enumerate}[label=\alph*)]
    \item The \orgshort{} shall not maintain a shadow database of all General Members. Communications to the full student body shall be facilitated through the Department Graduate Program Administrator.
    \item All information other than names and membership level is confidential and released only to the department staff upon specific request received in writing to \orgshort{}.
\end{enumerate}

\Sec{7. Removal of Membership}
\begin{enumerate}[label=\alph*)]
    \item A member may be removed for failure to comply with these bylaws or department policy, violation of the Student Code of Conduct, or violation of RPI Handbook of Student Rights \& Responsibilities.

    \item \textbf{Process:} (i) A written complaint is submitted to the \orgshort{} Council Membership; (ii) removal requires a two-thirds (2/3) vote of the \orgshort{} Council Membership; (iii) notice of decision and appeal options is provided in writing within five (5) business days.
\end{enumerate}

\Article{IV}{Elections}

\begin{enumerate}[label=\alph*)]
    \item \textbf{Timing:} Elections shall be held during the first week of November each year.
    \item Elections shall retain the following process:
    \begin{enumerate}[label=\roman*.]
        \item Graduate Program Administrator announces the election timeline at least two (2) weeks in advance;
        \item nominations open for one (1) week and may be self-nominations;
        \item voting occurs at a General Meeting or via secured online ballot open for 48 hours;
        \item simple majority elects;
        \item ties are resolved by a runoff between tied candidates.
    \end{enumerate}
    \item Students shall take office immediately upon announcement of election results, at which time the term of office for previous members shall expire.
    \item Any member who suspects unfair conduct shall notify the Graduate Program Administrator for further scrutiny.
    \item Vacancies on the Council and appointments to other bodies may be filled by appointment with the approval of two-thirds of the voting membership of the Council.
\end{enumerate}

\Article{V}{Officers}

\Sec{1. Officers and Duties}

\begin{enumerate}
    \item The Officers of the \orgshort{} shall consist of a President, a Secretary, and at least two (2) additional officers.
    
    \item The President (Mandatory) shall:
    \begin{enumerate}
        \item Act as the primary spokesperson for the Council.
        \item Preside over all Council meetings, or appoint a President Pro Tempore to do so if unable to attend.
        \item Set the agenda for General and Council meetings.
    \end{enumerate}

    \item The Secretary (Mandatory) shall:
    \begin{enumerate}
        \item Take minutes at all meetings and distribute them to Council members.
        \item Maintain the official records and membership roster of the \orgshort{}.
        \item Be responsible for advertising meetings and events of the Council.
    \end{enumerate}

    \item \textbf{Standard Roles for Reference} (The Council may elect members to these positions or define new ones as needed):
    \begin{enumerate}
        \item The Vice President shall act as President in their absence and coordinate operations not provided elsewhere.
        \item The Events Coordinator shall plan social, professional, and wellness events and coordinate logistics.
        \item The Publicity \& Communications Coordinator shall manage outreach (email, social media) and maintain the web presence.
    \end{enumerate}
\end{enumerate}

\Sec{2. Eligibility for Office}
\begin{enumerate}[label=\alph*)]
    \item Only elected Council Members are eligible to serve as Officers.
    \item No person may hold more than one Officer position concurrently.
    \item If an officer cannot complete their term, they shall inform the Council immediately.
\end{enumerate}

\Sec{3. Election of Officers}
\begin{enumerate}[label=\alph*)]
    \item \textbf{Timing:} Officer elections shall be held annually at the first meeting of the newly elected Council. This meeting must occur within two (2) weeks of the general Council results announcement.
    
    \item \textbf{Presiding Officer:} The outgoing Council President shall call this meeting and preside until the new President is elected. If the outgoing President is unavailable, the Graduate Program Administrator may preside.

    \item \textbf{Process:}
    \begin{enumerate}[label=(\roman*)]
        \item Nominations for President are taken from the floor (self-nominations permitted).
        \item Voting occurs via show of hands by the Council Members present.
        \item A two-thirds (2/3) majority of the Council Members is required to elect.
        \item Once elected, the new President immediately presides over the election of the Secretary and remaining officers.
    \end{enumerate}

    \item \textbf{Term:} Officers serve for the duration of their Council term (one year) or until their successors are elected.
\end{enumerate}

\Sec{4. Removal from Office}
\begin{enumerate}[label=\alph*)]
    \item An officer may be removed from their specific officer role (while retaining their Council seat) for failure to perform duties or violation of policies.
    \item \textbf{Process:} (i) A motion to remove an officer must be made by a current Council Member; (ii) the motion must be discussed at a meeting with at least five (5) days' notice; (iii) removal requires a two-thirds (2/3) vote of Council Members present and voting.
\end{enumerate}

\Sec{5. Succession and Vacancies}
\begin{enumerate}[label=\alph*)]
    \item In the event of a vacancy in the office of the President, the Vice President (if one exists) shall become President. If no Vice President exists, the Council shall hold an internal election to fill the position.
    \item Vacancies in other officer positions shall be filled by an internal election at the next regular Council meeting.
\end{enumerate}

\Article{VI}{Meetings}

\Sec{1. General Meetings}
\begin{enumerate}[label=\alph*)]
    \item The \orgshort{} shall hold a minimum of two (2) General Meetings each Fall and Spring semester.
    \item \textbf{Procedure \& Quorum:} Meetings follow an agenda set by the President and distributed at least 24 hours in advance. Quorum for business is fifty percent (50\%) of Council Members. Proxy or electronic participation may count toward quorum if allowed by the President. Actions pass by simple majority unless otherwise specified.
\end{enumerate}

\Article{VII}{Policies, Standing Rules, and By-Laws}

\Sec{1. Availability}
\begin{enumerate}[label=\alph*)]
    \item All policies, standing rules, and bylaws shall be available to any student of the department or interested member upon request and shall be published wherever these bylaws are published. These bylaws must be published on the \orgshort{} website, and on file with the \depname{} department administration.
\end{enumerate}

\Article{VIII}{Ratification and Amendments}

\Sec{1. Supremacy Clause}
If any part of these bylaws violates or contradicts the policies of Rensselaer, the Rensselaer Union, or applicable laws, the contradictory parts are invalid.

\Sec{2. Ratification}
\begin{enumerate}[label=\alph*)]
    \item These bylaws shall be effective immediately upon ratification by a two-thirds (2/3) majority of members with voting rights and subsequent approval of the \depname{} department administration.
    \item Amendments require a two-thirds (2/3) vote of members with voting rights present at the amendment meeting and subsequent approval by the \depname{} department administration.
\end{enumerate}

\Sec{3. Abrogation}
Upon adoption, these bylaws abrogates previous governing documents relating to the \orgshort{} without disrupting officers currently serving terms validly elected under a prior set of bylaws.

\end{document}
